% bibliography.tex — References
% Bibliography (book pages 726–730)

\begin{thebibliography}{76}

\bibitem{ablowitz1991}
Ablowitz, M.~J. and Clarkson, P.~A.
\textit{Solitons, Nonlinear Evolution Equations and Inverse Scattering}.
Cambridge: Cambridge University Press, 1991.

\bibitem{ablowitz1981}
Ablowitz, M.~J. and Segur, H.
\textit{Solitons and the Inverse Scattering Transform}.
Philadelphia: SIAM, 1981.

\bibitem{abramowitz1974}
Abramowitz, M. and Stegun, I.~A., eds.
\textit{Handbook of Mathematical Functions}.
New York: Dover, 1974.

\bibitem{acheson1990}
Acheson, D.~J.
\textit{Elementary Fluid Dynamics}.
Oxford: Oxford University Press, 1990.

\bibitem{aki1980}
Aki, K. and Richards, P.~G.
\textit{Quantitative Seismology}.
San Francisco: Freeman, 1980.

\bibitem{ames1992}
Ames, W.~F.
\textit{Numerical Methods for Partial Differential Equations}, 3rd ed.
New York: Academic Press, 1992.

\bibitem{antman1980}
Antman, S.~S.
The equations for large vibrations of strings.
\textit{Amer.\ Math.\ Monthly} \textbf{87}, 1980, pp.~359--370.

\bibitem{bender1999}
Bender, C.~M. and Orszag, S.~A.
\textit{Advanced Mathematical Methods for Scientists and Engineers}.
New York: Springer-Verlag, 1999.

\bibitem{benney1966}
Benney, D.~J.
Long waves on liquid films.
\textit{J.~Math.\ Phys.\ (Stud.\ Appl.\ Math.)} \textbf{45}, 1966, pp.~150--155.

\bibitem{benney1967}
Benney, D.~J. and Newell, A.~C.
The propagation of nonlinear wave envelopes.
\textit{J.~Math.\ Phys.\ (Stud.\ Appl.\ Math.)} \textbf{46}, 1967, pp.~133--139.

\bibitem{berg1966}
Berg, P.~W., and McGregor, J.~L.
\textit{Elementary Partial Differential Equations}.
San Francisco: Holden-Day, 1966.

\bibitem{bleistein1984}
Bleistein, N.
\textit{Mathematical Methods for Wave Phenomena}.
Orlando: Academic Press, 1984.

\bibitem{bleistein1986}
Bleistein, N. and Handelsman, R.~A.
\textit{Asymptotic Expansions of Integrals}.
New York: Dover, 1986.

\bibitem{boyce2000}
Boyce, W.~E. and Di~Prima, R.~C.
\textit{Elementary Differential Equations and Boundary Value Problems}, 7th ed.
New York: Wiley, 2000.

\bibitem{campbell1996}
Campbell, S.~L. and Haberman, R.
\textit{Introduction to Differential Equations with Boundary Value Problems}.
Boston: Houghton-Mifflin, 1996.

\bibitem{carrier1988}
Carrier, G. and Pearson, C.
\textit{Partial Differential Equations}, 2nd ed.
New York: Academic Press, 1988.

\bibitem{carslaw1959}
Carslaw, H.~S. and Jaeger, J.~C.
\textit{Conduction of Heat in Solids}, 2nd ed.
New York: Oxford University Press, 1959.

\bibitem{churchill1972}
Churchill, R.~V.
\textit{Operational Mathematics}, 3rd ed.
New York: McGraw-Hill, 1972.

\bibitem{churchill2000}
Churchill, R.~V., and Brown, J.~W.
\textit{Fourier Series and Boundary Value Problems}, 6th ed.
New York: McGraw-Hill, 2000.

\bibitem{courant1953}
Courant, R. and Hilbert, D.
\textit{Methods of Mathematical Physics}, Vols.~1 and 2.
New York: Wiley, 1953.

\bibitem{crc2002}
\textit{CRC Standard Mathematical Tables and Formulae}, 31th ed.
Boca Raton, FL: CRC Press, 2002.

\bibitem{dennemeyer1968}
Dennemeyer, R.
\textit{Introduction to Partial Differential Equations and Boundary Value Problems}.
New York: McGraw-Hill, 1968.

\bibitem{drazin1981}
Drazin, P.~G. and Reid, W.~H.
\textit{Hydrodynamic Stability}.
Cambridge: Cambridge University Press, 1981.

\bibitem{duff1966}
Duff, G.~F.~D. and Naylor, D.
\textit{Differential Equations of Applied Mathematics}.
New York: Wiley, 1966.

\bibitem{garabedian1998}
Garabedian, P.
\textit{Partial Differential Equations}, 2nd ed.
New York: Chelsea, 1998.

\bibitem{gardner1967}
Gardner, C.~S., Greene, J.~M., Kruskal, M.~D., and Miura, R.~M.
Method for solving the Korteweg-de Vries equation.
\textit{Phys.\ Rev.\ Lett.} \textbf{19}, 1967, pp.~1095--1097.

\bibitem{gelfand1955}
Gel'fand, I.~M. and Levitan, B.~M.
On the determination of a differential equation from its spectral function.
\textit{Amer.\ Math.\ Soc.\ Transl.}~1, Ser.~2, 1955, pp.~253--304.

\bibitem{glendenning1994}
Glendenning, P.
\textit{Stability, Instability, and Chaos}.
Cambridge: Cambridge University Press, 1994.

\bibitem{greenberg1978}
Greenberg, M.~D.
\textit{Foundations of Applied Mathematics}.
Englewood Cliffs, NJ: Prentice Hall, 1978.

\bibitem{guckenheimer1983}
Guckenheimer, J. and Holmes, P.
\textit{Nonlinear Oscillations, Dynamical Systems, and Bifurcation of Vector Fields}.
New York: Springer-Verlag, 1983.

\bibitem{haberman1998}
Haberman, R.
\textit{Mathematical Models: Mechanical Vibrations, Population Dynamics, and Traffic Flow}.
Philadelphia: Society of Industrial and Applied Mathematics, 1998.

\bibitem{hildebrand1976}
Hildebrand, F.~B.
\textit{Advanced Calculus for Applications}, 2nd ed.
Englewood Cliffs, NJ: Prentice Hall, 1976.

\bibitem{isaacson1994}
Isaacson, E. and Keller, H.~B.
\textit{Analysis of Numerical Methods}.
New York: Dover, 1994.

\bibitem{jackson1998}
Jackson, J.~D.
\textit{Classical Electrodynamics}, 3rd ed.
New York: Wiley, 1998.

\bibitem{kaplan1981}
Kaplan, W.
\textit{Advanced Mathematics for Engineers}.
Reading, MA: Addison-Wesley, 1981.

\bibitem{keener1988}
Keener, J.~P.
\textit{Principles of Applied Mathematics}.
Reading, MA: Addison-Wesley, 1988.

\bibitem{kevorkian2000}
Kevorkian, J.
\textit{Partial Differential Equations}, 2nd ed.
New York: Springer, 2000.

\bibitem{kevorkian1996}
Kevorkian, J. and Cole, J.~D.
\textit{Multiple Scale and Singular Perturbation Methods}.
New York: Springer-Verlag, 1996.

\bibitem{kinsman1984}
Kinsman, B.
\textit{Wind Waves}.
New York: Dover, 1984.

\bibitem{lax1968}
Lax, P.~D.
Integrals on nonlinear equations and solitary waves.
\textit{Comm.\ Pure Appl.\ Math.} \textbf{21}, 1968, pp.~467--490.

\bibitem{lighthill2002}
Lighthill, J.
\textit{Waves in Fluids}.
Cambridge: Cambridge University Press, 2002.

\bibitem{mathews1970}
Mathews, J., and Walker, R.~L.
\textit{Mathematical Methods of Physics}.
New York: W.~A. Benjamin, 1970.

\bibitem{mikhlin1967}
Mikhlin, S.~G., ed.
\textit{Linear Equations of Mathematical Physics}.
New York: Holt, Rinehart and Winston, 1967.

\bibitem{morse1953}
Morse, P.~M. and Feshbach, H.
\textit{Methods of Theoretical Physics}, Parts 1 and 2.
New York: McGraw-Hill, 1953.

\bibitem{morton1994}
Morton, K.~W. and Mayers, D.~F.
\textit{Numerical Solution of Partial Differential Equations}.
Cambridge: Cambridge University Press, 1994.

\bibitem{murray1993}
Murray, J.~D.
\textit{Mathematical Biology}, 2nd ed.
New York: Springer-Verlag, 1993.

\bibitem{nayfeh2002}
Nayfeh, A.~H.
\textit{Perturbation Methods}.
New York: Wiley, 2002.

\bibitem{newell1985}
Newell, A.~C.
\textit{Solitons in Mathematics and Physics}.
Philadelphia: Society for Industrial and Applied Mathematics, 1985.

\bibitem{newell1992}
Newell, A.~C. and Moloney, J.~V.
\textit{Nonlinear Optics}.
Redwood City, CA: Addison Wesley, 1992.

\bibitem{nicolis1995}
Nicolis, G.
\textit{Introduction to Nonlinear Science}.
Cambridge: Cambridge University Press, 1995.

\bibitem{powers1987}
Powers, D.~L.
\textit{Boundary Value Problems}, 3rd ed.
San Diego: Harcourt Brace Jovanovich, 1987.

\bibitem{prigogine1968}
Prigogine, I. and Lefever, R.
Symmetry-breaking instabilities in dissipative systems.
\textit{J.~Chem.\ Phys.} \textbf{48}, 1968, pp.~1695--1700.

\bibitem{protter1967}
Protter, M.~H. and Weinberger, H.~F.
\textit{Maximum Principles in Differential Equations}.
Englewood Cliffs, NJ: Prentice Hall, 1967.

\bibitem{roberts1966}
Roberts, G.~E. and Kaufman, H.
\textit{Tables of Laplace Transforms}.
New York: W.~B. Saunders Co., 1966.

\bibitem{sagan1989}
Sagan, H.
\textit{Boundary and Eigenvalue Problems in Mathematical Physics}.
New York: Dover, 1989.

\bibitem{scott1999}
Scott, A.
\textit{Nonlinear Science: Emergence and Dynamics of Coherent Structures}.
Oxford: Oxford University Press, 1999.

\bibitem{smith1986}
Smith, G.~D.
\textit{Numerical Solution of Partial Differential Equations}, 3rd ed.
New York: Oxford University Press, 1986.

\bibitem{stakgold1997}
Stakgold, I.
\textit{Green's Functions and Boundary Value Problems}, 2nd ed.
New York: Wiley, 1997.

\bibitem{strauss1992}
Strauss, W.~A.
\textit{Partial Differential Equations}.
New York: Wiley, 1992.

\bibitem{strang1986}
Strang, G.
\textit{Introduction to Applied Mathematics}.
Wellesley, MA: Wellesley-Cambridge Press, 1986.

\bibitem{strang1993}
Strang, G.
\textit{Introduction to Linear Algebra}.
Wellesley, MA: Wellesley-Cambridge Press, 1993.

\bibitem{street1973}
Street, R.~L.
\textit{The Analysis and Solution of Partial Differential Equations}.
Monterey, CA: Brooks/Cole, 1973.

\bibitem{strogatz1994}
Strogatz, S.~H.
\textit{Nonlinear Dynamics and Chaos}.
Reading, MA: Addison-Wesley, 1994.

\bibitem{turing1952}
Turing, A.
The chemical basis for morphogenesis.
\textit{Phil.\ Trans.\ Roy.\ Soc.\ London} \textbf{B237}, 1952, pp.~37--72.

\bibitem{verhulst1997}
Verhulst, F.
\textit{Nonlinear Differential Equations and Dynamical Systems}, 2nd ed.
New York: Springer-Verlag, 1997.

\bibitem{watson1995}
Watson, G.~N.
\textit{A Treatise on the Theory of Bessel Functions}, 2nd ed.
Cambridge: Cambridge University Press, 1995.

\bibitem{weinberger1995}
Weinberger, H.~F.
\textit{A First Course in Partial Differential Equations}.
New York: Dover, 1995.

\bibitem{whitham1999}
Whitham, G.~B.
\textit{Linear and Nonlinear Waves}.
New York: Wiley, reprint 1999.

\bibitem{winfree1987}
Winfree, A.~T.
\textit{When Time Breaks Down: The Three-Dimensional Dynamics of Electrochemical Waves and Cardiac Tissue}.
Princeton, NJ: Princeton University Press, 1987.

\bibitem{young1972}
Young, E.~C.
\textit{Partial Differential Equations}.
Boston: Allyn and Bacon, 1972.

\bibitem{yuen1978}
Yuen, H.~C. and Ferguson, W.~E.
Relationship between Benjamin-Feir instability and recurrence in the nonlinear Schrodinger equation.
\textit{Phys.\ Fluids} \textbf{21}, 1978, pp.~1275--1278.

\bibitem{yuen1975}
Yuen, H.~C. and Lake, B.~M.
Instabilities of waves on deep water.
\textit{Phys.\ Fluids} \textbf{18}, 1975, pp.~956--960.

\bibitem{zabusky1965}
Zabusky, N.~J. and Kruskal, M.~D.
Interaction of solitons in a collisionless plasma and the recurrence of initial states.
\textit{Phys.\ Rev.\ Lett.} \textbf{15}, 1965, pp.~240--243.

\bibitem{zachmanoglou1989}
Zachmanoglou, E.~C. and Thoe, D.~W.
\textit{Introduction to Partial Differential Equations with Applications}.
New York: Dover, 1989.

\bibitem{zauderer1989}
Zauderer, E.
\textit{Partial Differential Equations of Applied Mathematics}, 2nd ed.
New York: Wiley, 1989.

\bibitem{zwillinger1998}
Zwillinger, D.
\textit{Handbook of Differential Equations}, 3rd ed.
San Diego: Academic Press, 1998.

\end{thebibliography}
