% problems10.tex — Problem Set for Chapter 10

\begin{problemset}

\textbf{EMPIRICAL EXERCISES}

\medskip

(Skim Stock and Watson, 1993, Section~7 and Appendix~B, before answering.) The
file MPYR.ASC has annual data on the following:

\medskip
\noindent
Column 1: natural log of M1 (to be referred to as $m$)\\
Column 2: natural log of the NNP price deflator ($p$)\\
Column 3: natural log of NNP ($y$)\\
Column 4: the commercial paper rate in percent at an annual rate ($R$).

\medskip

The sample period is 1900 to 1989. This is the same data used by Stock and Watson
(1993) in their study of the U.S.\ money demand. See their Appendix~B for data
sources.

Questions~(a) and~(b) are for verifying the presumption that $m - p$ is cointe\-grated
with $(y, R)$. The rest of the questions are about estimating the cointegrating
vector.

\medskip

\begin{enumerate}[label=(\alph*)]
\item (Univariate unit-root tests)\quad Stock and Watson (1993, Appendix~B) report that
  the ADF $t^\mu$ and $t^\tau$ statistics, computed with 2 and 4 lagged first differences,
  fail to reject a unit root in each of $m - p$ and $R$ at the 10 percent level and that
  the unit root hypothesis is not rejected for $y$ with 4 lags, but is rejected at the
  10 percent (but not 5 percent) level with 2 lags. Verify these findings. In your
  tests, use the $t^\mu$ test for $R$, and $t^\tau$ for $m - p$ and $y$, and use asymptotic critical
  values (those critical values for $T = \infty$).

\item (Residual-based tests)\quad Conduct the Engle-Granger test of the null that $m - p$
  is not cointegrated with $y$ and $R$. Set $p = 1$ (which is what is selected by BIC).
  (If you do not include time in the regression, the ADF $t$-statistic derived from
  the residuals should be about $-4.7$. Case~2 discussed in Section~10.3 should
  apply. So the 5 percent critical value should be $-3.80$.)

\item (DOLS)\quad Reproduce the SOLS and DOLS estimates of the cointegrating vector
  reported in Table~10.2.

\item (Chow test)\quad Reproduce the DOLS estimates reported in Table~10.4. They are
  based on the following augmented cointegrating regression:
  \begin{align*}
    m_t - p_t
    &= \mu + \gamma_y y_t + \gamma_R R_t
      + \delta_0 D_t + \delta_y y_t D_t + \delta_R R_t D_t \\
    &\quad + \beta_{y0} \Delta y_t
      + \beta_{y,-1} \Delta y_{t+1} + \beta_{y,-2} \Delta y_{t+2}
      + \beta_{y1} \Delta y_{t-1} + \beta_{y2} \Delta y_{t-2} \\
    &\quad + \beta_{R0} \Delta R_t
      + \beta_{R,-1} \Delta R_{t+1} + \beta_{R,-2} \Delta R_{t+2}
      + \beta_{R1} \Delta R_{t-1} + \beta_{R2} \Delta R_{t-2}
      + v_t.
  \end{align*}
  The null hypothesis is that $\delta_0 = \delta_y = \delta_R = 0$. Set the $p$ in (10.4.16) to~2.
  Calculate the Wald statistic for the null hypothesis of no structural change.
\end{enumerate}

\end{problemset}
