% ch08.tex — Chapter 8: Directions for Further Study
\chapter{Directions for Further Study}\label{ch:further-study}

Although the results of the previous chapters cover a considerable range
of the possibilities of interest to economists, they also constitute but a
modest entry into the rich realm of modern econometrics. We may view
the territory before us as extending in a number of different though related directions. Among these are more general data generating processes
(DGPs); more general specifications for the models used to analyze these
DGPs; estimation procedures alternative to least squares and instrumental
variables; and consideration of the consequences and detection of model
misspecification.

\section{Extending the Data Generating Process}\label{sec:extending-dgp}

Loosely speaking, there are three dimensions in which the data generating
processes we consider can be extended: moments, memory, and heterogeneity. For example, although we consider unit root DGPs, we have not
discussed DGPs that involve deterministic time trends. Nevertheless, the
Markov law of large numbers (Theorem 3.7) or the McLeish law of large
numbers (Theorem 3.47) can be useful in establishing consistency in models with trending explanatory or instrumental variables. In fact, consistency
may happen ``faster'' in models with these variables because the error variance may become quite negligible in comparison to the magnitude of the
regression function $\bfX_t'\bfbeta_o$. Asymptotic normality can be established with
the help of the Lindeberg or Martingale-Lindeberg central limit theorems
(Theorems 5.6 or 5.24). In fact, conditions ensuring asymptotic normality in models with nonstochastic and possibly trending variables were the
subject of careful attention very early on (Grenander, 1954) and there is a
well-developed general theory now available (e.g., Crowder, 1980).

Deterministic time trends as usually considered constitute a growth in
the first moment of the dependent variable. Trends may occur in higher
moments as well. The key to handling such cases is to have available laws
of large numbers, central limit theorems and/or functional central limit theorems that permit such possibilities. For example, Wooldridge and White
(1988) give FCLT results for processes with possibly trending moments of
higher order.

We have considered DGPs satisfying the memory requirements of ergodicity and mixing. Recall also that we imposed the mixingale memory
requirement to establish the stationary ergodic CLT and FCLT. Although
it constitutes a restriction for stationary ergodic processes, the mixingale
notion can also be used to relax the memory conditions in the mixing context. Specifically, by considering stochastic processes that depend on an
infinite history of an underlying mixing process, but which depend primarily on the ``near epoch'' of the mixing process, we can obtain processes that
have longer memory than a simple mixing process but which inherit enough
of the properties of the underlying mixing process to satisfy the mixingale
condition. In fact, it can be shown that such ``near epoch dependent'' functions of a mixing process are mixingales that are sufficiently well behaved
so as to satisfy laws of large numbers, CLTs and FCLTs.

Such conditions were introduced by Ibragimov (1962) and treated by
Billingsley (1968). McLeish (1975a,b) used these conditions in developing
laws of large numbers, CLTs and FCLTs, which were introduced to econometrics by Gallant and White (1988). See Gallant and White (1988) and
Davidson (1994) for further discussion and details.

In Chapter 7, we considered integrated (unit root) processes, which exhibit considerably more dependence than mixing, mixingales, or ergodicity
permit. We see, however, that the large sample behavior of our estimators in these cases depends on the FCLT, whose validity is based on the
memory properties of the innovations to the integrated process. A memory
requirement intermediate to the mixingale and integrated processes considered here is that of fractional integration, a generalization of the notion of
integrated processes. A review of such processes is given by Sowell (1992)
and Baillie (1996). The tools required to study the large sample properties
of estimators associated with fractionally integrated DGPs are extensions
of the FCLT and the other tools of Chapter 7. With fractional integration,
the FCLT analogs deliver convergence to ``fractional Brownian motion.''
See, e.g., Taqqu (1975) for further discussion.

To keep our discussion of the FCLT in Chapter 7 relatively simple, we
impose the global covariance stationarity assumption, restricting the heterogeneity of the DGP more than is necessary to deliver the FCLT or
analogous results. See Wooldridge and White (1988), Davidson (1994), and
DeJong and Davidson (2000) for results that do not require global covariance stationarity.

\section{Nonlinear Models}\label{sec:nonlinear-models}

Throughout, we have restricted attention to models linear in the parameters, although we have allowed nonlinear restrictions among the parameters
to hold. A more general model that contains many situations of interest to
economists can be written as $\vec{q}_t(\bfX_t, \bfY_t, \bfbeta)$, where for some $\bfbeta_o$ it is assumed
that $\vec{q}_t(\bfX_t, \bfY_t, \bfbeta_o) = \bfeps_t$.

In the particular case we studied, the model has the form
%
\[
  \vec{q}_t(\bfX_t, \bfY_t, \bfbeta) = \bfY_t - \bfX_t'\bfbeta.
\]

Situations in which the elements of the dependent variable $\bfY_t$ take values
in a restricted subset of the real line are often of considerable interest. For
example, suppose $Y_t$ is a scalar that can take only the values 0 or 1 (the
``limited dependent variable'' case). In this case a relevant model is given
by
%
\[
  q(\bfX_t, Y_t, \bfbeta) = Y_t - F(\bfX_t'\bfbeta),
\]

where $F$ is a cumulative distribution function (e.g., the normal or logistic
c.d.f.). We interpret $F(\bfX_t'\bfbeta)$ as a model for $E(Y_t|\bfX_t) = P[Y_t = 1|\bfX_t]$.

There is a vast range of other possibilities of which the linear model is
merely a simple and convenient special case. Further, the nature of the
model adopted and the other knowledge available about the DGP can play
a determining role in the estimation procedures one employs.

\section{Other Estimation Techniques}\label{sec:other-estimation}

To estimate parameters of the model $\vec{q}_t(\bfX_t, \bfY_t, \bfbeta)$ just discussed, we can
employ the method of moments introduced and described briefly in Chapter
4. Specifically, if we have instrumental variables $\bfZ_t$ available such that
$E(\bfZ_t \bfeps_t) = \bfzero$, then we can attempt to estimate $\bfbeta_o$ by solving the problem
%
\[
  \min_{\bfbeta} \vec{q}(\bfbeta)'\bfZ\hat{\vec{P}}_n\bfZ'\vec{q}(\bfbeta),
\]

where $\vec{q}(\bfbeta)$ is the $np \times 1$ vector with $t$th block $\vec{q}_t(\bfX_t, \bfY_t, \bfbeta)$, so
%
\[
  \bfZ'\vec{q}(\bfbeta) = \sum_{t=1}^{n} \bfZ_t \vec{q}_t(\bfX_t, \bfY_t, \bfbeta).
\]

This corresponds to the method of moments procedure with $\vec{g}(\bfX_t, \bfY_t, \bfZ_t, \bfbeta)$
$= \bfZ_t \vec{q}_t(\bfX_t, \bfY_t, \bfbeta)$.

The properties of this method of moments estimator have been studied
by Amemiya (1977), Burguete, Gallant, and Souza (1982), and Hansen
(1982), among others.

To establish properties for estimators of nonlinear models analogous to
those obtained here, we need somewhat more powerful tools than those
given. In particular, repeated use is made of uniform laws of large numbers
and the mean value theorem for random functions (e.g., see Jennrich, 1969,
and White, 1994).

A leading alternative to the method of instrumental variables studied
here is the method of maximum likelihood. In fact, if one assumes that
the disturbances $\bfeps_t$ are independent and identically distributed as multivariate normal with unknown covariance matrix, then the optimal IV estimators of Chapter 4 can be shown to be asymptotically equivalent to the
maximum likelihood estimator under general conditions. There is a broad
range of situations where maximum likelihood and instrumental variables
are asymptotically equivalent (see Hausman, 1975, and Amemiya, 1977),
although this equivalence fails for the general case of nonlinear models previously mentioned. In that case, maximum likelihood can be shown to be
more efficient than instrumental variables (Amemiya, 1977).

Use of the method of maximum likelihood requires an assumption about
the distribution of the errors, whereas the instrumental variables method
does not. Thus, the method of instrumental variables is available in situations where a knowledge of the error distribution is absent or suspect.
Nevertheless, maximum likelihood estimation can be conducted as if the errors have the assumed distribution, whether this assumption is valid or not.
This procedure is known as quasi-maximum likelihood estimation, a member of the class of $M$-estimators (Huber, 1967), which contains a wealth
of useful and interesting estimators. By selecting an $M$-estimator appropriately, it is possible to obtain estimators that are robust to failure of
distributional assumptions or to certain plausible kinds of data errors.

Again, the study of these estimators requires, among other things, use
of uniform laws of large numbers and mean value theorems for random
functions. A general treatment of these estimators that also highlights the
parallels with IV estimators is that of Gallant and White (1988).

\section{Model Misspecification}\label{sec:model-misspecification}

Throughout this book, we have maintained the assumption that the data
generating relationship is known to be
%
\[
  \bfY_t = \bfX_t'\bfbeta_o + \bfeps_t, \quad t = 1, 2, \ldots\,.
\]

It would indeed be fortunate if the relationship between $\bfX_t$ and $\bfY_t$ were
ever truly ``known.'' Owing to the complexity of economic phenomena, it is
perhaps more realistic to suppose that the relationship between $\bfX_t$ and $\bfY_t$
is unknown. In this case, a linear relationship such as that just given can
be viewed as a convenient approximation but not necessarily as a definitive description of the relationship between $\bfX_t$ and $\bfY_t$. It then becomes
important to consider questions such as ``How is this approximation to be
interpreted?''; ``What are the properties of the estimated parameters of the
approximation?''; ``How can the approximation be improved?''; and ``How
can we tell if our approximation is exact?''.

For a discussion of these issues that builds on the material in this book
in a framework encompassing several of the extensions discussed in this
chapter, the reader is referred to \emph{Estimation, Inference, and Specification
Analysis} (White, 1994).

\section*{References}

\begin{description}

\item Amemiya, T. (1977). ``The maximum likelihood estimator and the nonlinear
  three-stage least squares estimator in the general nonlinear simultaneous equation model.'' \emph{Econometrica}, 45, 955--968.

\item Baillie, R. T. (1996). ``Long memory processes and fractional integration in econometrics.'' \emph{Journal of Econometrics}, 73, 5--59.

\item Billingsley, P. (1968). \emph{Convergence of Probability Measures}. Wiley, New York.

\item Burguete, J. F., A. R. Gallant, and G. Souza (1982). ``On unification of the
  asymptotic theory of nonlinear econometric models.'' \emph{Econometric Reviews}, 1,
  151--212.

\item Crowder, M. J. (1980). ``On the asymptotic properties of least-squares estimators
  in autoregression.'' \emph{Annals of Statistics}, 8, 132--146.

\item Davidson, J. (1994). \emph{Stochastic Limit Theory}. Oxford University Press, New York.

\item DeJong R. M. and J. Davidson (2000). ``The functional central limit theorem
  and weak convergence to stochastic integrals I: Weakly dependent processes,''
  forthcoming in \emph{Econometric Theory}, 16.

\item Gallant, A. R. and H. White (1988). \emph{A Unified Theory of Estimation and Inference for Nonlinear Dynamic Models}. Basil Blackwell, Oxford.

\item Grenander, U. (1954). ``On the estimation of regression coefficients in the case
  of an autocorrelated disturbance.'' \emph{Annals of Mathematical Statistics}, 25, 252--272.

\item Hansen, L. P. (1982). ``Large sample properties of generalized method of moments
  estimators.'' \emph{Econometrica}, 50, 1029--1054.

\item Hausman, J. A. (1975). ``An instrumental variable approach to full information
  estimators for linear and certain nonlinear structural models.'' \emph{Econometrica},
  43, 727--738.

\item Huber, P. J. (1967). ``The behavior of maximum likelihood estimates under nonstandard conditions.'' In \emph{Proceedings of the Fifth Berkeley Symposium on Mathematical Statistics and Probability}, vol.\ 1, pp.\ 221--233. University of California
  Press, Berkeley, California.

\item Ibragimov, I. A. (1962). ``Some limit theorems for stationary processes.'' \emph{Theory
  of Probability and Its Applications}, 7, 349--382.

\item Jennrich, R. I. (1969). ``Asymptotic properties of nonlinear least squares estimators.'' \emph{Annals of Mathematical Statistics}, 40, 633--643.

\item McLeish, D. L. (1975a). ``A maximal inequality and dependent strong laws.''
  \emph{Annals of Probability}, 3, 826--836.

\item[\phantom{x}] (1975b). ``Invariance principles for dependent variables.'' \emph{Zeitschrift f\"ur
  Wahrscheinlichkeitstheorie und Verwandte Gebiete}, 32, 165--178.

\item Sowell, F. (1992). ``Maximum-likelihood-estimation of stationary univariate fractionally integrated time-series models.'' \emph{Journal of Econometrics}, 53, 165--188.

\item Taqqu, M. S. (1975). ``Weak convergence to fractional Brownian motion and to
  the Rosenblatt process.'' \emph{Zeitschrift f\"ur Wahrscheinlichkeitstheorie und Verwandte Gebiete}, 31, 287--302.

\item White, H. (1994). \emph{Estimation, Inference and Specification Analysis}. Cambridge
  University Press, New York.

\item Wooldridge, J. M. and H. White (1988). ``Some invariance principles and central
  limit theorems for dependent heterogeneous processes.'' \emph{Econometric Theory},
  4, 210--230.

\end{description}
